---
title: "Mapping Oceanographic Data with R"
format: html
editor: visual
---

Goal: Plot spatially referenced point and polygon data on a basemap (google earth or bathymetry)

# Libraries we'll use

```{r}
library(tidyverse)
library(sf)
library(ggmap)
library(marmap)
```

# Importing data and formatting for spatial use

## Import cruise data
```{r}
oct <- read.csv("~/Desktop/Repos/ScallopRSA2021/data/OctCruiseData.csv") %>%
  rename(Lat = Latitude_degrees_start,
         Long = Longitude_degrees_start) %>%
  filter(Filtered == "",
         Cast == "Bottom") %>%
  mutate(Tow = ifelse(is.na(Tow_start_time), "N", "Y"))
```

## Convert coordinates from DM to DD
```{r}
chd <- substr(oct$Lat, 3, 3)[1]

# latitude
Lat_split_oct <- str_split_fixed(oct$Lat, chd, 2) %>%
                 as.data.frame()
Lat_split_oct$V2 <- str_remove_all(Lat_split_oct$V2, pattern = "'") %>%
                    as.numeric()
Lat_split_oct$V2 <- Lat_split_oct$V2/60
Lat_split_oct$V1 <- as.numeric(Lat_split_oct$V1)

oct$Lat <- Lat_split_oct$V1 + Lat_split_oct$V2
rm(Lat_split_oct)

# longitude
Long_split_oct <- str_split_fixed(oct$Long, chd, 2) %>%
                  as.data.frame()
Long_split_oct$V2 <- str_remove_all(Long_split_oct$V2, pattern = "'") %>%
                     as.numeric()
Long_split_oct$V2 <- Long_split_oct$V2/60
Long_split_oct$V1 <- as.numeric(Long_split_oct$V1)

oct$Long <- -(Long_split_oct$V1 + Long_split_oct$V2)
```

## Convert cruise data to a simple features object

"Simple Features" is a commonly used set of GIS standards for storing and accessing data representing predominantly two-dimensional geographic features (e.g. points, lines, polygons).

```{r}
# create geometry column in data frame from lat/long data
for(i in 1:nrow(oct)) {
  oct$geometry[i] <- st_point(x = c(oct$Long[i], oct$Lat[i])) %>%
    st_sfc(crs = 4326)
}

# create an sf object from data frame
station_sf <- st_sf(oct, crs = 4326)
```

# Build a basemap

## Use a satellite image

### Register Google API key
```{r, include=FALSE}
register_google(key = "AIzaSyBvvS_nvTml09h56OWd0Oniktayi0L8WDU")
```

```{r}
# extract satellite map from Google Maps
GB_map <- get_map(c(-68.5247, 40.8155), maptype = "satellite", zoom = 7, source = "google")

# create basemap using ggmap
satmap <- ggmap(GB_map) +
  coord_sf(crs = st_crs(4326)) +
  xlim(c(-70.5, -66)) +
  ylim(c(40, 42.5)) +
  theme(axis.title = element_blank())

satmap
```

## Use imported bathymetry
```{r}
# import bathymetric data from NOAA's server
GB_bathy <- getNOAA.bathy(lon1 = -74,
                          lon2 = -62,
                          lat1 = 37,
                          lat2 = 46,
                          resolution = 1)

# prepare data for mapping with ggplot
ggbathy <- GB_bathy %>%
  fortify() %>%
  mutate(depth_bins = cut(z,
                          breaks = c(0, -30, -55, -75, -90, -120, -150, -180,
                                     -780, -1380, -1980, -2580, -3180, -Inf)))

# create basemap
Georges <- ggplot() +
  geom_raster(data = ggbathy, aes(x, y, fill = depth_bins),
              interpolate = TRUE, alpha = 0.75) +
  coord_sf(xlim = c(-70.5, -66), ylim = c(40, 42.5), crs = st_crs(4326)) +
  theme_bw() +
  theme(axis.title = element_blank(),
        legend.position = "none")

Georges
```

## Change the depth scale to shades of blue
```{r}
Georges <- Georges +
  scale_fill_manual(values = c("#08306B", "#084184", "#08519C", "#1561A9",
                               "#2171B5", "#3282BE", "#4292C6", "#57A0CE",
                               "#6BAED6", "#85BCDC", "#9ECAE1", "#B2D3E8",
                               "#C6DBEF"))

Georges
```

## Use a higher-resolution coastline

GSHHG (Global Self-consistent, Hierarchical, High-resolution Geography) shoreline data can be downloaded from https://www.ngdc.noaa.gov/mgg/shorelines/data/gshhg/latest/.

Additional note: Spherical geometry is turned off because many polygons cannot be valid on S2 and R2 simultaneously, as outlined here: https://r-spatial.org/book/04-Spherical.html#validity-on-the-sphere

```{r}
# turn off spherical geometry to avoid "duplicate vertex" errors
sf_use_s2(FALSE)

mass <- read_sf("/Library/Frameworks/R.framework/Versions/4.2-arm64/Resources/library/gshhg-shp-2.3.7/GSHHS_shp/f/GSHHS_f_L1.shp") %>%
  st_make_valid() %>%
  st_crop(xmin = -71, xmax = -66, ymin = 40, ymax = 43)

Georges <- Georges +
  geom_sf(data = mass, fill = "darkgrey", colour = "black") +
  coord_sf(xlim = c(-70.5, -66), ylim = c(40, 42.5), crs = st_crs(4326))

Georges
```

# Plot points on a spatial grid

```{r}
Georges +
  geom_point(data = station_sf, aes(x = Long, y = Lat), shape = 17)
```

# Add polygons

## Import shapefile of scallop fishing management zones
```{r}
GB <- st_read("~/Desktop/Repos/ScallopRSA2021/data/2020SAMZones/GB_Estimation_Areas_2019_UTM19_PDT_SFModified.shp")
```

```{r}
Georges +
  geom_sf(data = GB, fill = NA, colour = "black",
          linewidth = 0.4, linetype = "dashed") +
  coord_sf(xlim = c(-70.5, -66), ylim = c(40, 42.5), crs = st_crs(4326))
```

## Back to the satellite map for fun
```{r}
# subset individual SAM zones
NLS_North <- subset(GB, NewSAMS == "NLS-North")
CL1_South <- subset(GB, NewSAMS == "CL1-South")
CL1_Sliver <- subset(GB, NewSAMS == "CL1-Sliver")
CL2_AccessSoutheast <- subset(GB, NewSAMS == "CL2-Access-Southeast")
SF <- subset(GB, NewSAMS == "SF")
CL2_North <- subset(GB, NewSAMS == "CL2-North")
CL1_Access <- subset(GB, NewSAMS == "CL1-Access")
NF <- subset(GB, NewSAMS == "NF")
CL2_Ext <- subset(GB, NewSAMS == "CL2-Ext")
GSC <- subset(GB, NewSAMS == "GSC")
NLS_SouthDeep <- subset(GB, NewSAMS == "NLS-South-Deep")
NLS_West <- subset(GB, NewSAMS == "NLS-West")
NLS_SouthShallow <- subset(GB, NewSAMS == "NLS-South-Shallow")
CL2_AccessSouthwest <- subset(GB, NewSAMS == "CL2-Access-Southwest")
SF_East <- subset(GB, NewSAMS == "SF-East")

# add to the map!
satmap <- satmap +
  geom_sf(data = NLS_North, fill = "palegreen2",
          col = "transparent", inherit.aes = FALSE) +
  geom_sf(data = CL1_South, fill = "firebrick3",
          col = "transparent", inherit.aes = FALSE) +
  geom_sf(data = CL1_Sliver, fill = "mediumorchid3",
          col = "transparent", inherit.aes = FALSE) +
  geom_sf(data = CL2_AccessSoutheast, fill = "lightskyblue",
          col = "transparent", inherit.aes = FALSE) +
  geom_sf(data = SF, fill = "darkorange2",
          col = "transparent", inherit.aes = FALSE) +
  geom_sf(data = CL2_North, fill = "seagreen2",
          col = "transparent", inherit.aes = FALSE) +
  geom_sf(data = CL1_Access, fill = "chartreuse3",
          col = "transparent", inherit.aes = FALSE) +
  geom_sf(data = NF, fill = "peachpuff2",
          col = "transparent", inherit.aes = FALSE) +
  geom_sf(data = CL2_Ext, fill = "slateblue1",
          col = "transparent", inherit.aes = FALSE) +
  geom_sf(data = GSC, fill = "steelblue1",
          col = "transparent", inherit.aes = FALSE) +
  geom_sf(data = NLS_SouthDeep, fill = "blueviolet",
          col = "transparent", inherit.aes = FALSE) +
  geom_sf(data = NLS_West, fill = "lightpink2",
          col = "transparent", inherit.aes = FALSE) +
  geom_sf(data = NLS_SouthShallow, fill = "tan3",
          col = "transparent", inherit.aes = FALSE) +
  geom_sf(data = CL2_AccessSouthwest, fill = "yellow2",
          col = "transparent", inherit.aes = FALSE) +
  geom_sf(data = SF_East, fill = "darkturquoise",
          col = "transparent", inherit.aes = FALSE) +
  coord_sf(crs = st_crs(4326))

satmap
```
## Add points on top
```{r}
satmap +
  geom_point(data = station_sf, aes(x = Long, y = Lat), shape = 17)
```

# Working with rasters (if time)
